---
title: "Assignment Markdown"

output:
  pdf_document: default
  html_document: default
---

```{r Load packages, include=FALSE}
#install packages if not already installed
if (!requireNamespace(c("ggplot2", "palmerpenguins", "janitor", "dplyr","ragg","svglite"), quietly = TRUE)) {
  install.packages(c("ggplot2", "palmerpenguins", "janitor", "dplyr","ragg","svglite"))
}

#load packages
library(palmerpenguins)
library(ggplot2)
library(janitor)
library(dplyr)
library(ragg)
library(svglite)

#call new functions from scripts
source("functions/cleaning.r") 
source("functions/plotting.r")
```


# **Question 1**

1a)

```{r, echo=FALSE}
# create a new object with cleaned data
clean.penguins <- penguins_raw %>% 
  clean_columns() %>% 
    shorten_species() 


# put the clean data into a csv file
write.csv(clean.penguins, "Data/clean.penguins.csv")

#filter the data 
flipper_data <- clean.penguins %>%
    subset_columns(c("flipper_length_mm", "species","sex")) %>%
    remove_NA()

#create a plot
flipper_violin<-ggplot(flipper_data, aes(x = species, y = flipper_length_mm)) +
  geom_violin() +
  labs(title="Graph showing differences in flipper morphology between species", x = "Species", y = "Flipper length", fill="Sex") +
  theme(panel.grid = element_blank())

flipper_violin

```
**Figure 1. A "bad" figure showing flipper length distribution by species in the Palmer penguins data set.**
 
 
1b) Write about how your design choices mislead the reader about the underlying data (200-300 words).
 
The graph does not include any units for the flipper length, meaning the readers have guess the units based on the flipper lengths they would expect to see in penguins. However this estimation is difficult as the genus of these species or the common name (penguins) is not stated anywhere in the graph.
 
Furthermore, the y axis is truncated, meaning it does not start at 0. This makes differences seem a lot bigger than they actually are as we have effectively zoomed into part of the graph and blown it out of proportion. 
 
The data points are not shown, which means we don't know how many data points are present in each of the species. Though the width of the violins show the density and consequently the distribution of flipper lengths, there is no axis or scale for the width of the violin. This means we don't know how many data points lie in each part of the violin. Overlaying the data points would make this more clear. Furthermore, without data points we cannot see whether there are any outliers and how strongly they influence the shape of the violin. 
 
The choice of a violin plot also makes it hard to compare distributions laterally, merely due to the gap between each of the violins. This is less influential in boxplots as boxplots only differ in the y axis due to the lack of curves. Finally, there are no grid lines, meaning it is hard to see where the median and quartiles lie on the y axis. 
 
 
**References:**
 
Molina, E., Viale, L., and Vázquez, P. (2022) "How should we design violin plots?," 2022 IEEE 4th Workshop on Visualization Guidelines in Research, Design, and Education (VisGuides). doi: 10.1109/VisGuides57787.2022.00006.
 
 
# **Question 2 - Data pipeline**
 
**Introduction.**
 
The current analysis investigates the characteristics of different species of penguin in the dataset in the Palmerpenguins package. Specifically, we will look at whether body mass varies between three species: Chinstrap penguins (Pygoscelis antarcticus), Gentoo penguins (Pygoscelis papua) and Adelie penguins (Pygoscelis adeliae). These three species all habit Antartica and are closely related. It may be interesting to see whether their diversification involves changes in body shape. We hypothesise that the mean body masses of the species are not all the same, with differences in body shape being a product of their diversification. 
 
Firstly, the raw data from the Palmer penguins data set was cleaned to make the column names uniform, concise, and both computer and human readable. The data set was filtered to only include body mass and species. 
 
```{r Data Processing and Cleaning, echo=TRUE}

#load the function definitions
source("functions/cleaning.r") 

#save the raw data as an object
write.csv(penguins_raw, "data/penguins_raw.csv")

#visualise the raw data
head(penguins_raw)

#create a new object with cleaned data
clean.penguins <- penguins_raw %>% 
    clean_columns() %>% 
  # function to remove capital letters from the column names, remove spaces in column names
  shorten_species() 
#function to shorten the species names to "Adelie", "Chinstrap", and "Gentoo"

#put the clean data into a csv file
write.csv(clean.penguins, "Data/clean.penguins.csv") 

#create a new object filtering out the variables that aren't needed
penguins.mass<- clean.penguins %>%
    subset_columns(c("body_mass_g", "species")) %>% #keep these variables
    remove_NA() #remove rows (samples) that have any NAs

#visualise the filtered object
head(penguins.mass)
```

 
A violin plot was constructed to show the distribution of body masses in the three species. This was done by creating a plot function, called from the script plotting.r present in the 'functions' folder of the project directory.
 
 
```{r Exploratory Plot, echo=TRUE, fig.height=5, fig.width=7, message=FALSE, warning=FALSE}
#plot data using a violin plot using function from plotting.r
exploratory.plot<-plot_function(penguins.mass) 
exploratory.plot
```
**Figure 2. Exploratory violin plot showing the distribution of body masses in three species of penguins in the Palmer penguins data set.** This plot shows that there is both intra- and inter-species variation of body mass in the data set. Though there is overlap in the distributions of Gentoo and Chinstrap, the widest parts of the violins (the medians) seem to be quite different. Statistical tests can be conducted to test the significance of the inter-species variation.
 
 
 
 
This plot was then saved as an .svg file as this file type has a vector format, which is resolution-independent and will thus will not lose resolution with changes in size. This can be found in the 'figures' folder in the project directory. 
 
```{r Save exploratory plot, message=FALSE, warning=FALSE}
#save the figure in vector format using function called from plotting.r
save_function(penguins.mass,
              "figures/save.plot.svg",
              size=size_inches,scaling=scaling)
```

  

`---`
 
**Statistical tests**
 
A one-way ANOVA statistical test was chosen to determine whether the difference in mean body masses between species was significant. ANOVA was used as it is a method of comparing the mean of a continuous variable across more than two categories. 
 
Null hypothesis (H0) : the mean body mass is equal across all three species.
Alternative hypothesis (H1) : the mean body mass is significantly different in at least one of the species. 
 

```{r ANOVA Test, echo=TRUE}
#fit a linear model to the data
penguins.lm<-lm(body_mass_g~species,penguins.mass)

#use ANOVA function 
anova.result<-anova(penguins.lm) # run ANOVA on the linear model
print(anova.result) # print ANOVA results table

```
**Figure 3. Figure showing the results of the one-way ANOVA test.** P<0.05, meaning there is sufficient evidence to reject the null hypothesis. This means that at least one of the species has a significantly different mean body mass than the others.

`---`

A post-hoc Tukey Kramer was conducted to determine which of the three species had a significant difference in body mass.

```{r Post hoc Tukey test, fig.height=5, fig.width=7}
#conduct post hoc Tukey Kramer test
tukey.test <- TukeyHSD(aov(penguins.lm))
print(tukey.test)

```
 
 
**Figure 4. Figure showing the results of a post-hoc Tukey Kramer test.** There is insignificant difference in the mean body mass of Chinstrap and Adelie, and a significant difference between Gentoo and Adelie, and Gentoo and Chinstrap.
 
`---`

**Discussion**
 
From these results, we ask the question of why Chinstrap and Adelie species are significantly lighter than  Gentoo. Such differences could be due to environmental effects, such as the availability of food in the different populations of species sampled. Higher food availability in the habitat of the Gentoo penguins could have led to this larger body mass. The differences could be a result of genetic divergence between species. If the latter is true, we might question whether smaller body mass is the ancestral state, given that two of the three species sampled have a similarly small body mass. 
 
It is interesting to note that the distribution of body masses in the three species differed. The Chinstrap sample had a fairly concentrated distribution around the mean, relative to the other two species that had a broader distribution. This might suggest that there is stronger stabilising selective pressure on the Chinstrap sample, causing a more uniform body mass within the sample. 
 
 
To conclude, we see a significantly larger body mass in the sample of Gentoo penguin species, compared to Chinstrap and Adelie samples, which had similarly small body masses. 



